Meta-analytic approaches and effect sizes to account for 'nuisance heterogeneity' in comparative physiology

### 2021-10-04 MS/ms.Rmd
Added reference to robust variance estimators
Changed wording a bit

Meta-analysis is a powerful tool to generate quantitatively informed answers to pressing global challenges. By distilling data from broad sets of research designs and study systems into standardized effect sizes, meta-analyses provide physiologists with opportunities to estimate overall effect sizes and understand the drivers of effect variability. Despite this ambition, research designs in the field of comparative physiology can appear, at the outset, as being vastly different to each other because of 'nuisance heterogeneity' (e.g., different temperatures or treatment dosages used across studies). Methodological differences across studies have led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with multilevel meta-regression models to both account for the factors driving differences across studies and make them more comparable. We assess the prevelance of nuisance hetrogeneity in the comparative physiology literature -- showing it is common and often not accounted for in analyses. We then formalise effect size measures (e.g., the temperature coefficient, $Q_{10}$) that provide comparative physiologists with a means to remove nuisance heterogeneity without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be applied to a variety of different contexts to derive new effects sizes and sampling variances opening new possibilities. By using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, comparative physiologists can explore exciting new questions and make results from large-scale data sets more accessible, comparable and widely interpretable.

@@

There are many phenomena driving effect heterogeneity. First, there is inherent uncertainty in estimating the ‘true’ population effect (e.g., a correlation coefficient) from smaller samples [@Borenstein2019]. A formal meta-analysis weights sample estimates by their precision, and removes this source of heterogeneity [@Gurevitch2018; @Koricheva2013]. Removing sampling variance is possible because we know, mathematically, how to calculate it for many effect measures. While unweighted analyses are common, weighting studies makes overall estimates more precise, and less susceptible to publication bias (see below), even if the overall mean itself is unbiased [@Morrissey2016a]. Second, there are sociological factors impacting heterogeneity. These include the ease with which novel and significant results are published relative to non-significant results, manifesting as publication bias [@Jennions2013; @Nakagawa2021b; @Rothstein2005]. Third, real biological processes can drive effect heterogeneity. This is particularly true in comparative physiology where we synthesise data from different species with varying life-histories, habitats, mating systems and reproductive modes. As comparative physiologists, this is the stuff that gets us up every morning! We can model these biological factors by including relevant predictors (moderator variables) and/or random effects (to account for non-independence of effects). By doing so, we can test predictions from hypotheses that we believe are key players impacting mean effect size and direction. This improves our understanding of the biological world around us and helps inform on future research directions.

@@

where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect (a random effect) for species *k* applied to effect size *i* [i.e., species effects because of shared evolutionary history, @Cinar2021]. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_{phylogeny}$ [i.e., *N*$(0, \sigma^{2}_{phylogeny}\textbf{A})$], where $\textbf{A}$ is a phylogenetic correlation matrix derived from a phylogenetic tree. $sp_{k[i]}$ is an additional species-specific effect for the *k*th species applied to effect size *i* [i.e., these are species-specific effects because of shared ecology and other factors, @Cinar2021], $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect size, $v_{i}$ (which is known as the sampling variance for the effect). This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is partially weighted by the inverse sampling variance of each effect size. Importantly, if we have effect sizes that are derived from the *same* sample of organisms (e.g., because traits are measured on the same sets of individuals), then we need to remove this dependency from the calculation of each $v_{i}$; we can then model the known sampling covariance between effect sizes by modifying $\textbf{I}$ to include off-diagonals that account for this covariance [See Box 1 and @Noble2017]. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$, see below for how it is defined) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially important sources of variation, such as differences in the methods applied, species differences, or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogeneity analysis' which we describe below. Importantly, this model provides us with the answer to our question: across all the species and studies manipulating egg T, how much of an overall effect does T have on offspring traits relative to control conditions?

>>>>

Meta-analysis is a powerful tool to generate quantitatively informed answers to pressing global challenges. By distilling data from broad sets of research designs and study systems into standardized effect sizes, meta-analyses provide physiologists with opportunities to estimate overall effect sizes and understand the drivers of effect variability. Despite this ambition, research designs in the field of comparative physiology can appear, at the outset, as being vastly different to each other because of 'nuisance heterogeneity' (e.g., different temperatures or treatment dosages used across studies). Methodological differences across studies have led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with multilevel meta-regression models to both account for the factors driving differences across studies and make them more comparable. We assess the prevelance of nuisance hetrogeneity in the comparative physiology literature -- showing it is common and often not accounted for in analyses. We then formalise effect size measures (e.g., the temperature coefficient, $Q_{10}$) that provide comparative physiologists with a means to remove nuisance heterogeneity without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be applied to a variety of different contexts to derive new effects sizes and sampling variances, opening up new possibilities for quantitative synthesis. By using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, comparative physiologists can explore exciting new questions and make results from large-scale data sets more accessible, comparable and widely interpretable.

@@

There are many phenomena driving effect heterogeneity. First, there is inherent uncertainty in estimating the ‘true’ population effect (e.g., a correlation coefficient) from smaller samples [@Borenstein2019]. A formal meta-analysis weights sample estimates by their precision, and removes this source of heterogeneity [@Gurevitch2018; @Koricheva2013]. Removing sampling variance is possible because we know, mathematically, how to calculate it for many effect measures. While unweighted analyses are common, weighting studies makes overall estimates more precise, and less susceptible to publication bias (see below), even if the overall mean itself is unbiased [@Morrissey2016a]. Second, there are sociological factors impacting heterogeneity. These include the ease with which novel and significant results are published relative to non-significant results, manifesting as publication bias [@Jennions2013; @Nakagawa2021b; @Rothstein2005]. Third, real biological processes can drive effect heterogeneity. This is particularly true in comparative physiology where we synthesise data from different species with varying life-histories, habitats, mating systems and reproductive modes. As comparative physiologists, this is the stuff that gets us up every morning! We can model these biological factors by including relevant predictors (moderator variables) and/or random effects (to account for non-independence of effects). By doing so, we can test predictions from hypotheses about the key players impacting mean effect size and direction. This improves our understanding of the biological world around us and helps inform on future research directions.

@@

where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect (a random effect) for species *k* applied to effect size *i* [i.e., species effects because of shared evolutionary history, @Cinar2021]. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_{phylogeny}$ [i.e., *N*$(0, \sigma^{2}_{phylogeny}\textbf{A})$], where $\textbf{A}$ is a phylogenetic correlation matrix derived from a phylogenetic tree. $sp_{k[i]}$ is an additional species-specific effect for the *k*th species applied to effect size *i* [i.e., these are species-specific effects because of shared ecology and other factors, @Cinar2021], $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect size, $v_{i}$ (which is known as the sampling variance for the effect). This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is partially weighted by the inverse sampling variance of each effect size. Importantly, if we have effect sizes that are derived from the *same* sample of organisms (e.g., because traits are measured on the same sets of individuals), then we need to remove this dependency from the calculation of each $v_{i}$; we can then model the known sampling covariance between effect sizes by modifying $\textbf{I}$ to include off-diagonals that account for this covariance [See Box 1 and @Noble2017]. Alternatively, robust variance estimators can also be used to deal with effect size non-independence [@Song2021; @Nakagawa2021]. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$, see below for how it is defined) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially important sources of variation, such as differences in the methods applied, species differences, or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogeneity analysis' which we describe below. Importantly, this model provides us with the answer to our question: across all the species and studies manipulating egg T, how much of an overall effect does T have on offspring traits relative to control conditions?
